---
title: "tidycorels"
output:
  github_document:
    toc: true
    toc_depth: 3
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## What are corels and tidycorels?

> Corels are ['Certifiably Optimal RulE ListS'](https://corels.eecs.harvard.edu/). They are short and simple human [interpretable rule lists](https://arxiv.org/pdf/1704.01701.pdf) created on categorical data.

The Corels authors [say](https://corels.eecs.harvard.edu/corels/whatarerulelists.html) that.. 
*"The advantage of rule lists as predictive models is that they are human interpretable, as opposed to black box models such as neural nets. Proponents of such non-interpretable models often claim that they are necessary to achieve greater accuracy, but we have shown that it is possible for rule lists to have comparable or even greater accuracy."*

`tidycorels::tidy_corels()` converts your dataframe into two text files in the format the R package [corels](https://cran.r-project.org/package=corels) expects. It returns the Corels rules as well as applying them to your dataframe using [`data.table::fifelse()`](https://rdatatable.gitlab.io/data.table/reference/fifelse.html) code. 
Your dataframe is also returned with fewer columns ready to be used in an insightful [alluvial](https://github.com/erblast/easyalluvial/blob/master/README.md) plot. It can reveal intuitively both how the Corels rules are applied to your dataframe, and where the classification is correct or incorrect. This reduced dataframe includes: the true label, the columns used in the Corels rules in rule order, then the label column the rules create. 

## Installation

```{r, eval=FALSE}
devtools::install_github("billster45/tidycorels")
```

## Simple example

Let's use tidycorels to classify cars in `datasets::mtcars` as either automatic or manual.

### Prepare dataframe for Corels

Using [`recipes`](https://recipes.tidymodels.org/) functions in [`tidymodels`](https://www.tidymodels.org/), columns with continuous values are binned or [discretised](https://recipes.tidymodels.org/reference/step_discretize.html) into categorical data. Each bin in each column is given its own 0/1 binary column using [`recipes::step_dummy()](https://recipes.tidymodels.org/reference/step_dummy.html). This is sometimes called one-hot encoding.

Corels also requires the label column is split into two columns representing each class. In this example, `am` is the label for Corels to classify where 0 = automatic and 1 = manual gears.

```{r}
library(tidymodels)
library(corels)
library(tidycorels)
library(easyalluvial)
library(parcats)
library(networkD3)
library(visNetwork)
library(formattable)

mtcars_recipe <-recipes::recipe(am ~ ., data = datasets::mtcars) %>%
  # 1 discretise continous variables into bins
  recipes::step_discretize(mpg, disp, hp, drat, wt, qsec, min_unique = 1) %>% 
  # 2 convert each value of each predictor into its own 0/1 binary column
  recipes::step_mutate_at(recipes::all_predictors(), fn = list(~ as.factor(.))) %>%
  recipes::step_dummy(recipes::all_predictors(), one_hot = TRUE) %>%
  # 3 convert each value of the outcome column into its own 0/1 binary column
  recipes::step_integer(recipes::all_outcomes(), zero_based = TRUE) %>% # ensure outcome is 0/1 rather than words
  recipes::step_mutate_at(recipes::all_outcomes(), fn = list(~ as.factor(.))) %>%
  recipes::step_dummy(recipes::all_outcomes(), one_hot = TRUE)

# Train the recipe
mtcars_recipe_prep <- recipes::prep(mtcars_recipe, 
                                    training = datasets::mtcars,
                                    retain = TRUE)

# Extract the pre-processed data
mtcars_preprocessed <- recipes::juice(mtcars_recipe_prep)
```

The prepared dataframe can now be used in the `tidycorels::tidy_corels()` function. We need to specify the names of the two columns that represent the label, and the delimiter that seperates the column name from the value each dummy column represents. The default seperator used by [`recipes::step_dummy()](https://recipes.tidymodels.org/reference/step_dummy.html) is an underscore.

All other arguments of `corels::corels()` are available to set, other than the following that are fixed by `tidycorels::tidy_corels()`: `rules_file` (generated from `df`), `labels_file` (generated from `df`), `log_dir` (set as `getwd()`), `verbosity_policy` (set as "minor").

```{r, results='hide'}
corels_mtcars <-
  tidycorels::tidy_corels(
    df = mtcars_preprocessed,
    label_cols = c("am_X0", "am_X1"),
    value_delim = "_",
    run_bfs = TRUE,
    calculate_size = TRUE,
    run_curiosity = TRUE,
    regularization = 0.01,
    curiosity_policy = 3,
    map_type = 1
  )
```

A list of useful objects is returned. First let's view the rules captured from the console output of `corels::corels()`.

```{r}
corels_mtcars$corels_console_output[4:8]
```

And see those rules converted to `data.table::fifelse()` code. 

```{r}
corels_mtcars$corels_rules_DT
```

And we can view the Corels rules as a [D3 network sankey diagram](https://christophergandrud.github.io/networkD3/#sankey) of the rules applied to the training data.

```{r}
networkD3::sankeyNetwork(# edges
                         Links = corels_mtcars$sankey_edges_df, 
                         Value = "value", 
                         Source = "source",
                         Target = "target", 
                         # nodes
                         Nodes = corels_mtcars$sankey_nodes_df, 
                         NodeID = "label",
                         # format
                         fontSize = 16, 
                         nodeWidth = 40,
                         sinksRight = TRUE
                         )
```

And with some manipualation of the nodes and edges data, the rules can be viewed as a [visNetwork](https://datastorm-open.github.io/visNetwork/layout.html) visualisation.

```{r}
rule_count <- base::nrow(corels_mtcars$rule_performance_df)
# extract the rule order (or levels)
level <- corels_mtcars$rule_performance_df %>% 
  dplyr::mutate(level = dplyr::row_number()) %>% 
  dplyr::rename(level_label = rule) %>% 
  dplyr::select(level_label,level)

# rename the edges
edges <- corels_mtcars$sankey_edges_df %>% 
  dplyr::rename(from = source, to = target) %>% 
  dplyr::mutate(title = value)

# add the levels
nodes <- corels_mtcars$sankey_nodes_df %>% 
  dplyr::rename(id = ID) %>% 
  dplyr::mutate(level_label = stringi::stri_trim_both(stringi::stri_sub(label,1,-4))) %>% 
  dplyr::left_join(level) %>% 
  dplyr::mutate(level = case_when(is.na(level) ~ as.numeric(rule_count),TRUE ~as.numeric(level))) %>% 
  dplyr::rename(title = level_label)

visNetwork::visNetwork(nodes, edges, width = "100%") %>% 
  visNetwork::visNodes(size = 12) %>% 
  visNetwork::visEdges(arrows = "to") %>% 
  visNetwork::visHierarchicalLayout(direction = "UD", 
                                    levelSeparation = 80) %>% 
  visNetwork::visInteraction(navigationButtons = TRUE) %>% 
  visNetwork::visOptions(highlightNearest = list(enabled = T, hover = T), 
                         nodesIdSelection = T,
                         collapse = TRUE)
```

### Alluvial plot

A dataframe of just the true label, the columns used in the Corels rules, and the Corels label from those rules is also available. The columns have been ordered in the order of the Corels rules so that they work well in an [alluvial](https://github.com/erblast/easyalluvial/blob/master/README.md) plot.

```{r}
p <- corels_mtcars$alluvial_df %>%
  easyalluvial::alluvial_wide(stratum_width = 0.2,
                              NA_label = "not used") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "Corels if-then-else logic",
    subtitle = " From truth (far left column) to Corels classification (far right column)"
  )
p
```

The leftmost column is the true label for each car, either automatic (0) or manual (1). The rightmost column is the Corels classification label after applying the rules from left to right. 

The alluvial plot clearly shows which rules are important in the classifcation and exactly how the classifiation is arrived at. For example, if we follow one green path of manual cars (am_X1 = 1), they have more than 3 gears (gear_X3 = 0), low weight (wt_bin_1 ==1). The shape of the engine (vs_x0) column is not used.

We can also create an [interactive](https://erblast.github.io/easyalluvial/articles/parcats.html) version of the alluvial plot.

```{r}
parcats::parcats(p = p,
                 data_input = corels_mtcars$alluvial_df,
                 marginal_histograms = FALSE,
                 hoveron = 'dimension',
                 hoverinfo = 'count',
                 labelfont = list(size = 11)
)
```

### Performance on training data

We can also create a confusion matrix to examine the performance of the rules.

```{r}
conf_matrix <-
  corels_mtcars$alluvial_df %>%
  yardstick::conf_mat(
    truth = "am_X1",
    estimate = "corels_label"
  )

ggplot2::autoplot(conf_matrix, "heatmap")
```

### Performance of each rule

A data frame of the performance of each rule is also provided.

```{r}
corels_mtcars$rule_performance_df %>% 
  dplyr::mutate(rule_perc_correct = formattable::color_tile("white", "orange")(rule_perc_correct)) %>%
  dplyr::mutate(rule_fire_count = formattable::color_tile("white", "lightblue")(rule_fire_count)) %>%
  kableExtra::kable(escape = F,
                   caption = "Corels Performance for each rule in order") %>%
  kableExtra::kable_styling("hover", full_width = F)
```

### Performance on new data

Use `tidycorels::predict_corels()` to apply the Corels rules to a new dataframe (e.g. test data). It also returns the smaller data frame intended for an [alluvial](https://github.com/erblast/easyalluvial/blob/master/README.md) plot. Examples can be found in the [Articles](https://billster45.github.io/tidycorels/articles/) section.
